The role of cilia-mediated cerebrospinal fluid flow in intraventricular transport in zebrafish embryos

### 2025-02-03 paper/main.tex
Edits in Methods - Part I

@@

\section*{Methods}\label{sec:methods}
\subsection*{Zebrafish maintenance and strains in photoconversion experiments}
The animal facilities of zebrafish~(\emph{danio rerio}) we used are approved by the Norwegian Food Safety Authority (NFSA, Mattilsynet). Furthermore, the zebrafish were maintained in accordance with the guidelines set by the NFSA and the European Communities Council Directive.

The larval and adult zebrafish were raised under standard husbandry conditions at 28.5°C in a Techniplast Zebtech Multilinking system. The fish tanks were kept at constant pH 7.0 and 685 $\mu$S with a 14/10 hr light/dark cycle. From fertilization to 3 dpf, larvae were maintained in egg water (1.2 g marine salt and 0.1\% methylene blue in 20 L reverse osmosis (RO) water) and subsequently transferred to artificial fish water (AFW) (1.2 g marine salt in 20 L RO water). The zebrafish lines used in the experiments were \emph{ccdc103(dnaaf19)\textsuperscript{tn222a}} (\emph{schmalhans (smh)})~\cite{Jau-NianChen1997Left-rightZebrafish}(received from Drummond lab, MGH) and \emph{Tg(gfap:Gal4FF)\textsuperscript{nw7Tg}}~\cite{DiazVerdugo2019Glia-neuronSeizures}, \emph{Tg(5xuas:Signal-Dendra2)\textsuperscript{nw21Tg}}. Animals were in the pigmentless \emph{nacre\textsuperscript{b692}}(\emph{mitfa\textsuperscript{-/-}})~\cite{JamesA.Lister1999NacreFate} background.

Wholemount~\emph{in vivo} live imaging and photoconversion experiments were performed with zebrafish larvae at the 2 days post-fertilization (dpf) stage obtained from inbreeding of heterozygous \emph{smh\textsuperscript{+/-};Tg(gfap:Gal4FF);Tg(5xuas:Signal-Dendra2}) adult animals, as described in D’Gama~\emph{et al.}~\cite{DGama2024Cilia-mediatedBrain}. Controls consisted of either wild-type (\emph{smh}\textsuperscript{+/+}) or \emph{smh}\textsuperscript{+/-} heterozygous zebrafish from the same breeding. The mutants were identified based on their curved body~\cite{Jau-NianChen1997Left-rightZebrafish}.

@@

\subsection*{Generation of \emph{Tg(5xuas:Signal-Dendra2)\textsuperscript{nw21Tg}} line}
The open reading frame (ORF) of the neuropeptide y (npy) signal peptide (\emph{npy}: ENSDARG00000036222, Q1LW93) was fused with the N-terminal of zebrafish codon optimized Dendra2 DNA sequence. Then the conjugated DNA sequence was synthesized with EcoR1 enzyme sites at the 5’ and 3’ ends (GenScript Biotech), and the synthesized signal-Dendra2 DNA was inserted into the pT2MUASMCS vector~\cite{Asakawa2008GeneticZebrafish} through restriction enzyme cloning (GenScript Biotech).

To generate the transgenic line, the 5xuas:Signal-Dendra2 plasmid DNA was used. A volume of 2 nl of a mixture of the plasmid DNA (60 pg) and tol2 mRNA (10 pg) was microinjected into one-cell stage embryos, as described in Jeong~\emph{et al.}~\cite{Jeong2024TheZebrafish}. The injected embryos were raised to adulthood (F0). Germline-transmitted founder zebrafish were identified by breeding with multiple Gal4 transgenic lines. Stable F1 embryos expressing the Gal4-driven Dendra2 signals were screened and raised to adult zebrafish.

@@

The time-lapse images were acquired in a single plane, covering the telencephalic, diencephalic and rhombencephalic ventricles simultaneously, with a frequency of 0.37--0.38 Hz (2.60--2.68 sec / frame). The size of the images was $1024\times400$ pixels. In total, 300 images were acquired per fish. The first 10 images were scanned without photoconversion to obtain a baseline value of fluorescence intensity, and the next 290 images were scanned while performing photoconversion.

For photoconversion, ‘Bleaching’ and ‘Region’ in the ZEN software were used. A 405 nm laser was focused into a circular area (diameter: 16 $\mu$m, scan 0.77 $\mu$sec/pixel) of the dorsoposterior diencephalic ventricle with 100\% laser power. The laser was illuminated in the area repeatedly between each scanning. After imaging, the fish health was checked and only data from healthy fish were analyzed.
Data were collected and analyzed from three separate experiments, with a total of 8 controls and 6 mutants.

\subsection*{Photoconversion data analysis}
All data analysis was performed in MATLAB and the codes are available in the repository of this article: \url{repo-link}.

Following import of the image data, the photoconverted channel was aligned to correct for drift in $x,y$ directions using previously developed codes~\cite{Reiten2017Motile-Cilia-MediatedComputations, Ringers2023NovelEpithelia}. Only stable recordings were further analyzed. 
To increase the signal to noise ratio, the images were downsampled by a block average using a resampling factor of 6 (using the MATLAB function blockproc). The change of fluorescence $($$(F-F_0)/F_0$, hereafter referred to as $\Delta F$$)$ was calculated for each pixel at each time point of the data. Here, the value $F_0$ is the average fluorescence intensity for the baseline (the first 10 time-lapse images) acquired before photoconversion.

@@

Anesthetized 2 dpf control zebrafish were injected with 1 nl of a mixture containing 0.1\% w/v fluorescent beads (SPHERO Fluorescent Yellow Particles 1\% w/v, $F = 0.16$ mm) diluted in 7.5 mg/ml 70 kDa rhodamine B isothiocyanate-dextran (RITC-dextran; Sigma-Aldrich, R9379) dissolved in artificial cerebrospinal fluid. Artificial cerebrospinal fluid composition was as follows: 124 mM NaCl, 22 mM D-(+)-Glucose, 2.0 mM KCl, 1.6 mM MgSO\textsubscript{4} $\cdot$ 7 H\textsubscript{2}O, 1.3 mM KH\textsubscript{2}PO\textsubscript{4}, 24 mM NaHCO\textsubscript{3}, 2.0 mM CaCl\textsubscript{2} $\cdot$ 2 H\textsubscript{2}O. The needles used for the injections were pulled with a Sutter Instrument Co. Model P-2000 from thin-walled glass capillaries (1.00 mm; VWR) and cut open with a forceps. A volume of 1 nl of the solution was injected with a pressure injector (Eppendorf Femtojet 4i) in the rostral rhombencephalic ventricle. The pressure and time were calibrated for each needle using a 0.01 mm calibration slide.

Following injection, the zebrafish was directly placed under the confocal microscope and 1200 images were acquired at a frequency of 13.1 Hz. A single optical section was obtained. Particle tracking was done using TrackMate~\cite{Tinevez2017TrackMate:Tracking} in Fiji/ImageJ~\cite{Schindelin2012Fiji:Analysis} and plotted using MATLAB, as described in Jeong~\emph{et al.}~\cite{Jeong2022MeasurementTelencephalon}. The parameters for TrackMate were as follows. The DoG detector was used with the settings: threshold: 2.0, with median filtering; radius: 2.0, with subpixel localization. Next, the Simple LAP tracker was used with settings: max frame gap: 2; alternative linking cost factor: 1.05; linking max distance: 5.0; gap closing max distance: 2.0; splitting max distance: 15.0; allow gap closing: true; allow track splitting: false; allow track merging: false; merging max distance: 15.0; cutoff percentile: 0.9. Only the particle tracks with at least 30 data points, while simultaneously covering a distance of at least 8.8 $\mu$m, were plotted in~\cref{fig:figure1} (panels \textbf{b} and \textbf{c}).

\subsection*{Image-based computational geometry of zebrafish brain ventricles}
We generated a 3D representation of the brain ventricular walls using confocal imaging data of a zebrafish embryo injected intraventricularly with a 70 kDa dye at 2 dpf~\cite{Olstad2019CiliaryDevelopment}. From the imaging data, we constructed a computational mesh of the ventricles (\cref{fig:figure1}, panel \textbf{d}). The interior volume was meshed with fTetWild~\cite{Hu2020FastWild}. The standard mesh consisted of 132 134 tetrahedral cells associated with a total of 943 820 degrees of freedom, and the mesh had a maximal (minimal) edge length of 14.9 $\mu$m (4.68 $\mu$m). At 2 dpf, the ventricular system consists mainly of 3 cavities: the telencephalic  (anterior), the di-/mesencephalic  (middle) and the rhombencephalic (posterior) ventricles, which are connected by ducts.

Motile cilia were shown to line certain parts of the ventricular surface~\cite{Olstad2019CiliaryDevelopment}, as marked with pink in~\cref{fig:figure1} (panel \textbf{d}). Beating motion of the cilia generates cerebrospinal fluid (CSF) flow with rotational structures in the brain ventricles, resulting in a compartmentalized flow~(\cref{fig:figure1}, panels \textbf{a}--\textbf{c}). Meanwhile, the beating heart of the zebrafish embryo induces a pulsatile CSF flow component.
\begin{figure}[hbt!]
    \centering
    \includegraphics[width=\linewidth]{graphics/figure1.png}
    \caption{\lyng{Experimenting with velocity colorbars, thoughts on the two options? Or have two separate as it is now?}}
    \label{fig:figure1}
\end{figure}

\subsection*{Computational model of cerebrospinal fluid dynamics}
We model the flow of cerebrospinal fluid (CSF) in the brain ventricles by the time-dependent, incompressible Stokes equations: find the fluid velocity $\uu = \uu(\xx, t) = (u_x(\xx, t), u_y(\xx, t), u_z(\xx, t))$ and the fluid pressure $p = p(\xx, t)$ for $\xx \in \Omega$ and time $t>0$, such that
\begin{subequations}
    \begin{align}
  \rho\pdifft{\uu} - \nabla \cdot \bsig &= \mathbf{0}\mbox{ in }\Omega,\label{eq:stokes_eq_mom}\\
  \nabla \cdot \uu &= 0\mbox{ in }\Omega,
  \end{align}
  \label{eq:stokes_eqs}% 
\end{subequations}%
where the stress tensor is defined as $\bsig(\uu, p) = 2\mu\beps(\uu) - p\mathbf{I}$ and $\beps(\uu) = \frac{1}{2}\left(\nabla \uu + (\nabla\uu)^T\right)$ is the strain-rate tensor. Bold-face characters denote vectors or tensors, and $\mathbf{I}$ is the identity tensor in three dimensions. Furthermore, $\rho$ is the CSF density and $\mu$ is the dynamic viscosity of the CSF. We introduce the tangential traction $\bsigpar$ as the tangential component of the traction $\hat{\bsig}=\bsig\nn$:
\begin{equation*}
    \bsigpar = P_{\nn}(\hat{\bsig}) = (\mathbf{I} - \nn\otimes\nn)\hat{\bsig},
\end{equation*}
where $\nn$ is the outer unit normal of the surface $\Gamma = \partial\Omega$, and $P_{\nn}(\rr)$ is the tangential  projection of a vector $\rr$ onto $\Gamma$. We consider three types of boundary conditions:  (i) \emph{tangential traction} (no flow normal to the boundary $\uu \cdot \nn = 0$ with a force $\btau$ applied tangentially on the boundary $\bsigpar = \btau$), (ii) \emph{normal pressure} (a pressure $\Tilde{p}$ applied in the normal direction: $(\mu\nabla\uu - p\mathbf{I})\nn = \Tilde{p}(t)\nn$), and (iii) \emph{free slip} (no flow normal to the boundary $\uu \cdot \nn = 0$ and no tangential forces $\bsigpar = {\bf 0}$). Unless noted otherwise, the free-slip condition is applied on the boundary. We use the initial conditions $\uu(\xx, 0)=\mathbf{0}$ and $p(\xx, t) = 0$ for all $\xx\in\Omega$.

\subsection*{Cilia-driven flow}
On the boundary $\Gc$ of the ciliated regions of the ventricle walls~(pink-colored facets in~\cref{fig:figure1}, panel \textbf{d}), we impose the tangential traction $\bsigpar=\btau$ to represent the forces from the cilia acting on the cerebrospinal fluid. We set $\btau(\xx, \rr) = \tau \lambda(\xx) P_{\nn}(\rr)$, where $\lambda(\xx)$ is a function that is chosen differently for the cilia regions considered~(see~\cref{subsec:appendixA1}). The sign of the vector $\rr=\pm(1, 0, 1)$ also depends on the cilia region. Both $\rr$ and the value $\tau=6.5\times 10^{-4}$ Pa were chosen by comparing simulations with experimental data~\cite{Olstad2019CiliaryDevelopment}.

\subsection*{Flow induced by cardiac pulsations}\label{subsec:pressure_bcs}
To induce a pulsatile flow inside the brain ventricles, we set $\Tilde{p}(t)=0$ and $\Tilde{p}(t)=-A\sin{\omega t}$ on two separate parts of the boundary $\Gp$, located anterior and posterior, respectively, on the brain ventricles geometry~(blue-colored facets in~\cref{fig:figure1}, panel \textbf{d}). The amplitude $A=1.5\times 10^{-3}$ Pa determines the magnitude of the pulsatile flow, and was calibrated based on experimental data~\cite{Olstad2019CiliaryDevelopment}. The angular cardiac frequency $\omega=2\pi f=6.97$ rad/s was determined by measurements of the cardiac frequency $f=2.22$ Hz.

\iffalse
\subsection*{Computational cerebrospinal fluid dynamics in a moving reference frame}
 An Arbitrary Lagrangian Eulerian (ALE) formulation~\cite{san2009convergence} is used as an alternative to the normal pressure boundary conditions to model back-and-forth flow induced by cardiac pulsations. In a reference frame moving with a velocity $\uuref=(A_{\mathrm{ref}}\sin{\omega t}, 0, 0)$, we reformulate the Stokes equations~\eqref{eq:stokes_eqs} using an ALE formulation:
\begin{subequations}
    \begin{alignat}{2}
        \rho\bigg(\pdifft{\uu} + (\uu_{\mathrm{ref}}\cdot\nabla)\uu\bigg) &= \nabla\cdot\bm{\sigma}(\uu, p), \label{eq:mom_eq_ALE}\\
        \nabla\cdot\uu &= 0.
    \end{alignat}
    \label{eq:stokes_eqs_ALE}%
\end{subequations}%
We then solve for the velocity $\mathbf{w} = \uu - \uuref$ in the moving reference frame, where $\uu$ is the velocity in the fixed reference frame. The amplitude $A_{\mathrm{ref}}=919$ $\mu$m/s, calibrated based on experimental data, determines the magnitude of the pulsatile flow, and $\omega=2\pi f$ is the angular cardiac frequency. In the ALE formulation, we consider the impermeability boundary condition $\mathbf{w}\cdot\nn=0$ in the normal direction and the free-slip boundary condition $\bsigpar(\mathbf{w}, p) = \mathbf{0}$ in the tangential direction.
\fi

@@

%-----DISCRETIZATION AND MODEL IMPLEMENTATION-----%
\subsection*{Numerical approximation of the flow equations}
The Stokes equations~\eqref{eq:stokes_eqs} were discretized with a finite element method. In time, we employ a first-order implicit Euler discretization. We chose the timestep size $\Delta t = 0.02252$ s such that 20 timesteps represent one cardiac cycle. For the spatial discretization, we used piecewise linear Brezzi-Douglas-Marini (BDM) elements for the velocity. The BDM elements have degrees of freedom associated with integral moments of the velocity component normal to the facets of elements~\cite{Brezzi1985TwoProblems}, thus we enforced the impermeability condition $\uu\cdot\nn=0$ strongly. We imposed the boundary conditions for both the tangential traction $\bsigpar$ and the normal pressure $\Tilde{p}(t)$ weakly. For the pressure, we used piecewise constant elements, referred to as zeroth order discontinuous Galerkin (DG) elements. The element combination $\mathrm{BDM}_1$--$\mathrm{DG}_{0}$ yields a stable discretization scheme for the Stokes equations~\cite{Stenberg1989SomeEquations}. Since the divergence of the velocity is a subspace of the pressure space for this scheme, the divergence-free condition of the velocity is satisfied exactly on each element~\cite{Boffi2008FiniteProblem}, ensuring mass conservation. The BDM discretization includes a penalty parameter $\gamma$ which needs to be chosen large enough to ensure stability of the finite element method. We set $\gamma=10.$

>>>>

\newcommand{\fixme}[1]{\textcolor{red}{#1}}

@@

\section*{Methods}
\label{sec:methods}

\mer{@HH: I suggest you write \cref{fig:figure1}\textbf{a}, rather than \cref{fig:figure1} panel \textbf{a}, throughout. Increases signal-to-noise ratio.}

\subsection*{Zebrafish maintenance and strains in photoconversion experiments}
The zebrafish~(\emph{danio rerio}) animal facilities used are approved by the Norwegian Food Safety Authority (NFSA, Mattilsynet). Furthermore, the zebrafish were maintained in accordance with the guidelines set by the NFSA and the European Communities Council Directive. The larval and adult zebrafish were raised under standard husbandry conditions at 28.5°C in a Techniplast Zebtech Multilinking system. The fish tanks were kept at constant pH 7.0 and 685 $\mu$S with a 14/10 hr light/dark cycle. From fertilization to 3 dpf, larvae were maintained in egg water (1.2 g marine salt and 0.1\% methylene blue in 20 L reverse osmosis (RO) water) and subsequently transferred to artificial fish water (AFW) (1.2 g marine salt in 20 L RO water). The zebrafish lines used in the experiments were \emph{ccdc103(dnaaf19)\textsuperscript{tn222a}} (\emph{schmalhans (smh)})~\cite{Jau-NianChen1997Left-rightZebrafish}(received from Drummond lab, MGH) and \emph{Tg(gfap:Gal4FF)\textsuperscript{nw7Tg}}~\cite{DiazVerdugo2019Glia-neuronSeizures}, \emph{Tg(5xuas:Signal-Dendra2)\textsuperscript{nw21Tg}}. Animals were in the pigmentless \emph{nacre\textsuperscript{b692}}(\emph{mitfa\textsuperscript{-/-}})~\cite{JamesA.Lister1999NacreFate} background.

Wholemount~\emph{in vivo} live imaging and photoconversion experiments were performed with zebrafish larvae at the 2 days post-fertilization (dpf) stage obtained from inbreeding of heterozygous \emph{smh\textsuperscript{+/-};Tg(gfap:Gal4FF);Tg(5xuas:Signal-Dendra2}) adult animals, as described by D’Gama~\emph{et al.}~\cite{DGama2024Cilia-mediatedBrain}. Controls consisted of either wild-type (\emph{smh}\textsuperscript{+/+}) or \emph{smh}\textsuperscript{+/-} heterozygous zebrafish from the same breeding. The mutants were identified based on their curved body~\cite{Jau-NianChen1997Left-rightZebrafish}.

@@

\subsection*{Generation of \emph{Tg(5xuas:Signal-Dendra2)\textsuperscript{nw21Tg}} line}
\mer{MER: Could this heading be made understandable by non-experts?}
The open reading frame (ORF) of the neuropeptide y (npy) signal peptide (\emph{npy}: ENSDARG00000036222, Q1LW93) was fused with the N-terminal of zebrafish codon optimized Dendra2 DNA sequence. Next the conjugated DNA sequence was synthesized with EcoR1 enzyme sites at the 5’ and 3’ ends (GenScript Biotech), and the synthesized signal-Dendra2 DNA was inserted into the pT2MUASMCS vector~\cite{Asakawa2008GeneticZebrafish} through restriction enzyme cloning (GenScript Biotech).  To generate the transgenic line, the 5xuas:Signal-Dendra2 plasmid DNA was used. A volume of 2 nl of a mixture of the plasmid DNA (60 pg) and tol2 mRNA (10 pg) was microinjected into one-cell stage embryos, as described by Jeong~\emph{et al.}~\cite{Jeong2024TheZebrafish}. The injected embryos were raised to adulthood (F0). Germline-transmitted founder zebrafish were identified by breeding with multiple Gal4 transgenic lines. Stable F1 embryos expressing the Gal4-driven Dendra2 signals were screened and raised to adult zebrafish.

@@

The time-lapse images were acquired in a single plane, covering the telencephalic, diencephalic and rhombencephalic ventricles simultaneously, with a frequency of 0.37--0.38 Hz (2.60--2.68 sec / frame). The size of the images was $1024\times400$ pixels. In total, 300 images were acquired per fish. The first 10 images were scanned without photoconversion to obtain a baseline value of fluorescence intensity, and the next 290 images were scanned while performing photoconversion. For photoconversion, ‘Bleaching’ and ‘Region’ in the ZEN software were used. A 405 nm laser was focused into a circular area (diameter: 16 $\mu$m, scan 0.77 $\mu$sec/pixel) of the dorsoposterior diencephalic ventricle with 100\% laser power. The laser was illuminated in the area repeatedly between each scanning. After imaging, the fish health was checked and only data from healthy fish were analyzed.
Data were collected and analyzed from three separate experiments, with a total of 8 controls and 6 mutants.

\subsection*{Photoconversion data analysis}

Photoconversion data analysis was performed in MATLAB and the codes are \fixme{openly available\cite{zenodo-archive-for-repo-link-fixme}}. Following import of the image data, the photoconverted channel was aligned to correct for drift in $x,y$ directions using previously developed codes~\cite{Reiten2017Motile-Cilia-MediatedComputations, Ringers2023NovelEpithelia}. Only stable recordings were further analyzed. To increase the signal-to-noise ratio, the images were downsampled by a block average using a resampling factor of 6 (using the MATLAB function blockproc). The change of fluorescence $($$(F-F_0)/F_0$, hereafter referred to as $\Delta F$$)$ was calculated for each pixel at each time point of the data. Here, the value $F_0$ is the average fluorescence intensity for the baseline (the first 10 time-lapse images) acquired before photoconversion.

@@

Anesthetized 2 dpf control zebrafish were injected with 1 nl of a mixture containing 0.1\% w/v fluorescent beads (SPHERO Fluorescent Yellow Particles 1\% w/v, $F = 0.16$ mm) diluted in 7.5 mg/ml 70 kDa rhodamine B isothiocyanate-dextran (RITC-dextran; Sigma-Aldrich, R9379) dissolved in artificial CSF. Artificial CSF composition was as follows: 124 mM NaCl, 22 mM D-(+)-Glucose, 2.0 mM KCl, 1.6 mM MgSO\textsubscript{4} $\cdot$ 7 H\textsubscript{2}O, 1.3 mM KH\textsubscript{2}PO\textsubscript{4}, 24 mM NaHCO\textsubscript{3}, 2.0 mM CaCl\textsubscript{2} $\cdot$ 2 H\textsubscript{2}O. The needles used for the injections were pulled with a Sutter Instrument Co. Model P-2000 from thin-walled glass capillaries (1.00 mm; VWR) and cut open with a forceps. A volume of 1 nl of the solution was injected with a pressure injector (Eppendorf Femtojet 4i) in the rostral rhombencephalic ventricle. The pressure and time were calibrated for each needle using a 0.01 mm calibration slide.

Following injection, the zebrafish was directly placed under the confocal microscope and 1200 images were acquired at a frequency of 13.1 Hz. A single optical section was obtained. Particle tracking was done using TrackMate~\cite{Tinevez2017TrackMate:Tracking} in Fiji/ImageJ~\cite{Schindelin2012Fiji:Analysis} and plotted using MATLAB, as described by Jeong~\emph{et al.}~\cite{Jeong2022MeasurementTelencephalon}. The parameters for TrackMate were as follows. The DoG detector was used with the settings: threshold: 2.0, with median filtering; radius: 2.0, with subpixel localization. Next, the Simple LAP tracker was used with settings: max frame gap: 2; alternative linking cost factor: 1.05; linking max distance: 5.0; gap closing max distance: 2.0; splitting max distance: 15.0; allow gap closing: true; allow track splitting: false; allow track merging: false; merging max distance: 15.0; cutoff percentile: 0.9. Only the particle tracks with at least 30 data points, while simultaneously covering a distance of at least 8.8 $\mu$m, were shown (\cref{fig:figure1}, panels \textbf{b} and \textbf{c}).

\subsection*{Image-based computational geometry of zebrafish brain ventricles}
We generated a 3D representation of the brain ventricular walls using confocal imaging data of a zebrafish embryo injected intraventricularly with a 70 kDa dye at 2 dpf~\cite{Olstad2019CiliaryDevelopment}. From the imaging data, we constructed a computational mesh of the ventricles (\cref{fig:figure1}, panel \textbf{d}). The interior volume was meshed with fTetWild~\cite{Hu2020FastWild}. The standard mesh consisted of 132 134 tetrahedral cells and had a maximal (minimal) \mer{edge length} of 14.9 $\mu$m (4.68 $\mu$m). At 2 dpf, the ventricular system consists mainly of 3 cavities: the telencephalic  (anterior), the di-/mesencephalic  (middle) and the rhombencephalic (posterior) ventricles, which are connected by ducts. Parts of the ventricular surfaces were marked to be lined with motile cilia~\cite{Olstad2019CiliaryDevelopment} (\cref{fig:figure1}\textbf{d}, pink markers).
\mer{MER: @HH: Check if you mean edge length or mesh cell size here.}
\begin{figure}%[hbt!]
    \centering
    \includegraphics[width=\linewidth]{graphics/figure1.png}
    \caption{\lyng{Experimenting with velocity colorbars, thoughts on the two options? Or have two separate as it is now?}}
    \label{fig:figure1}
\end{figure}

\subsection*{Computational CSF dynamics}

The beating motion of the cilia generates CSF flow with steady rotational structures~\cite{Olstad2019CiliaryDevelopment} within the brain ventricles (\cref{fig:figure1}, panels \textbf{a}--\textbf{c}). In addition, cardiac pulsations induces pulsatile CSF flow. We model this flow of CSF in the brain ventricles by the time-dependent, incompressible Stokes equations which read as follows. Find the CSF velocity $\uu = \uu(\xx, t) = (u_x(\xx, t), u_y(\xx, t), u_z(\xx, t))$ and the CSF pressure $p = p(\xx, t)$ for $\xx \in \Omega$ and time $t>0$, such that
\begin{subequations}
    \begin{align}
  \rho\pdifft{\uu} - \nabla \cdot \bsig &= \mathbf{0}\mbox{ in }\Omega,\label{eq:stokes_eq_mom}\\
  \nabla \cdot \uu &= 0\mbox{ in }\Omega,
  \end{align}
  \label{eq:stokes_eqs}% 
\end{subequations}%
where the stress tensor is defined as $\bsig(\uu, p) = 2\mu\beps(\uu) - p\mathbf{I}$ and $\beps(\uu) = \frac{1}{2}\left(\nabla \uu + (\nabla\uu)^T\right)$ is the strain-rate tensor. Bold-face characters denote vectors or tensors, and $\mathbf{I}$ is the identity tensor in three dimensions. Furthermore, $\rho$ is the CSF density and $\mu$ is the dynamic viscosity of the CSF. We introduce the tangential traction $\bsigpar$ as the tangential component of the traction $\hat{\bsig}=\bsig\nn$:
\begin{equation*}
    \bsigpar = P_{\nn}(\hat{\bsig}) = (\mathbf{I} - \nn\otimes\nn)\hat{\bsig},
\end{equation*}
where $\nn$ is the outer unit normal of the boundary surface $\Gamma = \partial\Omega$, and $P_{\nn}(\rr)$ is the tangential  projection of a vector $\rr$ onto $\Gamma$. We consider three types of boundary conditions:  (i) \emph{tangential traction} (no flow normal to the boundary $\uu \cdot \nn = 0$ with a force $\btau$ applied tangentially on the boundary $\bsigpar = \btau$), (ii) \emph{normal pressure} (a pressure $\Tilde{p}$ applied in the normal direction: $(\mu\nabla\uu - p\mathbf{I})\nn = \Tilde{p}(t)\nn$), and (iii) \emph{free slip} (no flow normal to the boundary $\uu \cdot \nn = 0$ and no tangential forces $\bsigpar = {\bf 0}$). Unless noted otherwise, the free-slip condition is applied on the boundary. We use the initial condition\mer{s} $\uu(\xx, 0)=\mathbf{0}$ \mer{and $p(\xx, t) = 0$} for all $\xx\in\Omega$. \mer{MER: I don't think you use an initial condition on the pressure. Please check and remove if so. You can also just write: ``We assume that the system starts at rest ($\uu(t = 0) = \mathbf{0}$)''.}

\subsection*{Cilia-driven flow}
On the ciliated regions of the ventricle walls $\Gc$ (\cref{fig:figure1}, panel \textbf{d}), we impose a constant-in-time tangential traction $\bsigpar=\btau$ to represent the net forces of the cilia acting on the CSF. We set $\btau(\xx, \rr) = \tau \lambda(\xx) P_{\nn}(\rr)$, where the function $\lambda(\xx)$ and the sign of the vector $\rr=\pm(1, 0, 1)$ varies depending on the cilia region (see~\cref{subsec:appendixA1}). Both $\rr$ and the value $\tau=6.5\times 10^{-4}$ Pa were chosen by model calibration against the experimental data~\cite{Olstad2019CiliaryDevelopment}.

\subsection*{Flow induced by cardiac pulsations}\label{subsec:pressure_bcs}
To drive a cardiac-induced pulsatile flow inside the brain ventricles, we set $\Tilde{p}(t)=0$ and $\Tilde{p}(t)=-A\sin{\omega t}$ on anterior and posterior parts of the boundary $\Gp$~(\cref{fig:figure1}, panel \textbf{d}, blue markers). The amplitude $A=1.5\times 10^{-3}$ Pa determines the magnitude of the pulsatile flow, and was calibrated based on experimental data~\cite{Olstad2019CiliaryDevelopment}. The angular cardiac frequency $\omega=2\pi f=6.97$ rad/s was set based on a measured cardiac frequency of $f=2.22$ Hz.

%% \iffalse
%% \subsection*{Computational cerebrospinal fluid dynamics in a moving reference frame}
%%  An Arbitrary Lagrangian Eulerian (ALE) formulation~\cite{san2009convergence} is used as an alternative to the normal pressure boundary conditions to model back-and-forth flow induced by cardiac pulsations. In a reference frame moving with a velocity $\uuref=(A_{\mathrm{ref}}\sin{\omega t}, 0, 0)$, we reformulate the Stokes equations~\eqref{eq:stokes_eqs} using an ALE formulation:
%% \begin{subequations}
%%     \begin{alignat}{2}
%%         \rho\bigg(\pdifft{\uu} + (\uu_{\mathrm{ref}}\cdot\nabla)\uu\bigg) &= \nabla\cdot\bm{\sigma}(\uu, p), \label{eq:mom_eq_ALE}\\
%%         \nabla\cdot\uu &= 0.
%%     \end{alignat}
%%     \label{eq:stokes_eqs_ALE}%
%% \end{subequations}%
%% We then solve for the velocity $\mathbf{w} = \uu - \uuref$ in the moving reference frame, where $\uu$ is the velocity in the fixed reference frame. The amplitude $A_{\mathrm{ref}}=919$ $\mu$m/s, calibrated based on experimental data, determines the magnitude of the pulsatile flow, and $\omega=2\pi f$ is the angular cardiac frequency. In the ALE formulation, we consider the impermeability boundary condition $\mathbf{w}\cdot\nn=0$ in the normal direction and the free-slip boundary condition $\bsigpar(\mathbf{w}, p) = \mathbf{0}$ in the tangential direction.
%% \fi

@@

%-----DISCRETIZATION AND MODEL IMPLEMENTATION-----%
\subsection*{Numerical approximation of the flow equations}

\mer{associated with a total of 943 820 degrees of freedom}

The Stokes equations~\eqref{eq:stokes_eqs} were discretized with a finite element method. In time, we employ a first-order implicit Euler discretization. We chose the timestep size $\Delta t = 0.02252$ s such that 20 timesteps represent one cardiac cycle. For the spatial discretization, we used piecewise linear Brezzi-Douglas-Marini (BDM) elements for the velocity. The BDM elements have degrees of freedom associated with integral moments of the velocity component normal to the facets of elements~\cite{Brezzi1985TwoProblems}, thus we enforced the impermeability condition $\uu\cdot\nn=0$ strongly. We imposed the boundary conditions for both the tangential traction $\bsigpar$ and the normal pressure $\Tilde{p}(t)$ weakly. For the pressure, we used piecewise constant elements, referred to as zeroth order discontinuous Galerkin (DG) elements. The element combination $\mathrm{BDM}_1$--$\mathrm{DG}_{0}$ yields a stable discretization scheme for the Stokes equations~\cite{Stenberg1989SomeEquations}. Since the divergence of the velocity is a subspace of the pressure space for this scheme, the divergence-free condition of the velocity is satisfied exactly on each element~\cite{Boffi2008FiniteProblem}, ensuring mass conservation. The BDM discretization includes a penalty parameter $\gamma$ which needs to be chosen large enough to ensure stability of the finite element method. We set $\gamma=10.$
